---
title: "Study of the efficacy of Berenil applied to trypanosomosis's infected cattle"
author: "Gerard Castellà, Mathieu Marauri and Lluis Ramon"
date: May 20th, 2015
output: 
  beamer_presentation:
    theme: Berlin
    colortheme: "default"
    fonttheme: "default"
---

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library("knitr")
library("ggplot2")
library("xtable")
library("tidyr")
library("dplyr")
library("nlme")
opts_chunk$set(echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE)
options("xtable.comment" = FALSE)
```

```{r}
cows <- read.table("data/cattle_mes dades.txt", header = TRUE, 
                   sep = "\t", dec = ",", na.strings = "")

names(cows) <- c("id", "dose", "pcv", "time", "nbirth")
cows$dose <- factor(cows$dose, levels = c("L", "M", "H"))
cows$pcv.b <- as.numeric(cows$pcv > 20)
cows$pcv.f <- factor(cows$pcv > 20, labels = c("Unhealthy", "Healthy"))
cows$time.f <- paste("Time", cows$time)
cows.com <- na.omit(cows)
cows.com$idDose <- as.factor(paste(cows.com$id, cows.com$dose, sep = "_"))
```


# Introduction

The trypanosomosis's parasite is responsible for a disease that affects cows.

Effects of the disease:

* anemia
* loss of appetite
* death

# Objective

> __Assess the efficacy of different doses of Berenil in infected cattle.__


# Dataset

* 10 cows
* Response variable: Healthy / Unhealthy
* Covariates: _dose_, _time_, _nbirth_
* 3 dose levels (High, Medium, Low)

# Outcome categorization

* Literature review (healthy cow PCV 24-46)
* Practical Modeling

```{r outcome_categorization, results='asis'}
table_pcv_time <- table(cows$pcv.f , cows$time.f, useNA = "ifany")
rownames(table_pcv_time)[3] <- "Missing value"
taula <- xtable(table_pcv_time, digits=0)
print(taula, table.placement = "H")
````

# Assumptions

```{r}
library("ggplot2")
library("scales")
library("grid")
arrows <- data.frame(x = 1, xend = 17, y = c(6, 10, 14), yend = c(6, 10, 14))
df <- data.frame(x = c(0, 18), y = c(0, 18))
garrows <- ggplot(aes(x, y), data = df) + geom_segment(aes(x = x, y = y, xend = xend, yend = yend), arrow = arrow(length = unit(0.3,"cm")), data = arrows) + ylim(c(0, 18)) + coord_cartesian(xlim = c(-2, 18))

xdots <- c(2:4, 8:10, 12:14)
x3dots <- c(xdots, xdots, xdots)
ydots <- rep(c(6, 10, 14), each = 9)
dose1 <- rep(c("Low", "Medium", "High"), each = 3)
dose2 <- rep(c("Medium", "Low", "High"), each = 3)
dose3 <- rep(c( "High", "Low", "Medium"), each = 3)

timeIntervals <- data.frame(x = x3dots, y = ydots, dose = c(dose1, dose2, dose3))
timeIntervals$dose <- factor(timeIntervals$dose , levels = c("Low", "Medium", "High"))
g2 <- garrows + geom_point(aes(x = x, y = y, colour = dose), data = timeIntervals, size = 3)

segments <- data.frame(x = x3dots, xend = x3dots + 1, y = ydots, yend = ydots, dose = c(dose1, dose2, dose3))
segments$dose <- factor(segments$dose , levels = c("Low", "Medium", "High"))
segments <- segments[segments$x %in% c(2, 3, 8, 9, 12, 13), ]
g3 <- g2 + geom_segment(aes(x = x, y = y, xend = xend, yend = yend, colour = dose), data = segments, size = 1)
# g3 <- g3 + ylab("") + xlab("time") + theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), plot.background = element_rect(colour = "#DCDCDC"))
g3 <- g3 + ylab("") + xlab("time") + theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())
g4 <- g3 +  scale_x_continuous(breaks = xdots,  labels = c(1:3, 1:3, 1:3))

g4 <- g4 + theme_bw() + theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())
  
library("png")
cow <- readPNG("images/cow.png")
cowRed <- readPNG("images/cowRed.png")
cowBlue <- readPNG("images/cowBlue.png")
cowGreen <- readPNG("images/cowGreen.png")

cowA1 <-  annotation_raster(cow, xmin = -1.3, xmax = -1.3 + 1.58, ymin = 5.4, ymax = 6.6)
cowA2 <-  annotation_raster(cow, xmin = -1.3, xmax = -1.3 + 1.58, ymin = 9.4, ymax = 10.6)
cowA3 <-  annotation_raster(cow, xmin = -1.3, xmax = -1.3 + 1.58, ymin = 13.4, ymax = 14.6)

gCows <- g4 + cowA1 + cowA2 + cowA3

cowT11 <-  annotation_raster(cowRed, xmin = 2.34, xmax = 2.34 + 1.32, ymin = 6.5, ymax = 7.5)
cowT12 <-  annotation_raster(cowGreen, xmin = 2.34, xmax = 2.34 + 1.32, ymin = 10.5, ymax = 11.5)
cowT13 <-  annotation_raster(cowBlue, xmin = 2.34, xmax = 2.34 + 1.32, ymin = 14.5, ymax = 15.5)

cowT21 <-  annotation_raster(cowGreen, xmin = 8.34, xmax = 8.34 + 1.32, ymin = 6.5, ymax = 7.5)
cowT22 <-  annotation_raster(cowRed, xmin = 8.34, xmax = 8.34 + 1.32, ymin = 10.5, ymax = 11.5)
cowT23 <-  annotation_raster(cowRed, xmin = 8.34, xmax = 8.34 + 1.32, ymin = 14.5, ymax = 15.5)

cowT31 <-  annotation_raster(cowBlue, xmin = 12.34, xmax = 12.34 + 1.32, ymin = 6.5, ymax = 7.5)
cowT32 <-  annotation_raster(cowBlue, xmin = 12.34, xmax = 12.34 + 1.32, ymin = 10.5, ymax = 11.5)
cowT33 <-  annotation_raster(cowGreen, xmin = 12.34, xmax = 12.34 + 1.32, ymin = 14.5, ymax = 15.5)

gDef <- gCows + cowT11 + cowT12 + cowT13 +
   cowT21 + cowT22 + cowT23 +
   cowT31 + cowT32 + cowT33

gDef + scale_color_discrete(guide = 'none') + theme(text = element_text(size = 25))
```


# Statistical methods

* Generalized Estimating Equations (GEE) 
* Generalized Linear Mixed Model (GLMM)
* Missing Data Analysis

# Generalized Estimating Equations


# Generalized Linear Mixed Model

\begin{table}[H]
\centering
\begin{tabular}{rrrrr}
  \hline
 & df & AIC & p-value \\ 
  \hline
$\beta_{0} + + (\beta_{1} + b_{1i})time + \beta_2 doseMedium + \beta_3 doseHigh$ & 6 & 36.592  &  \\ 
 $\beta_{0} +  + (\beta_{1} + b_{1i})time + \beta_2 doseMedium + \beta_3 doseHigh + \beta_4 nbirth$ & 7 & 38.335 & 0.6123 \\ 
 $\beta_{0} +  (\beta_{1} + b_{1i})time + \beta_2 doseMedium + \beta_3 doseHigh + \beta_4 dose*time$ & 8 & 38.998 & 0.4506 \\  
   \hline
\end{tabular}
\caption{Anova table comparing the different models with the one with fixed effects for time and dose.} 
\label{anovas:Random_slope}
\end{table}

# Generalized Linear Mixed Model

Limitations of the results.

* Small number of observations,
* complete separation.

# Missing Data Analysis


# Conclusion

* The high dose is much more efficient.
* No difference between medium and low dose.
* The number of births has a relative negative impact.

# Limitations

* Information could be added to the data set.
* No control group.
* Small sample. Bias by selection?
* Missing values could be further analysed.



--------



> Thank you for your attention! 
